---
title: "MODULE 1: What Changed? - Identifying DEGs in Small Cell Lung Cancer"
author: "mkutmon"
date: "December 2025"
output: html_document
---

# 🎯 Module Goal
Identify which genes are significantly changed in lung cancer compared to healthy controls.
The dataset used is available on GEO: https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE60052

---

# SETUP

## Install & Load Packages

```{r packages, message=FALSE, warning=FALSE}
# Load packages
library(EnhancedVolcano)
library(rstudioapi)
library(readxl)

cat("✓ Packages loaded successfully!\n")
```

## Set Working Directory

```{r}
# Set working directory to script location
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

# Create output folder
out.folder <- "output/"
if(!dir.exists(out.folder)) dir.create(out.folder)

cat("✓ Working directory:", getwd(), "\n")
cat("✓ Output folder created\n")
```

---

# LOAD DATA

**What's in this dataset:**
The dataset contains gene expression measurements comparing small cell lung cancer sample with healthy control lung tissue.

**Columns:**
- **GeneID**: Ensembl gene identifier
- **GeneName**: Gene name (e.g., CDKN1A, TP53)
- **log2FC**: Log2 fold change (Lung Cancer vs. healthy)
  - **Positive** = UP in lung cancer
  - **Negative** = DOWN in lung cancer
- **pvalue**: P-value for this comparison
- **padj**: correction for multiple testing


```{r load-data}
# Load lung cancer dataset
data <- read_excel("lungcancer-dataset.xlsx")

# Quick check
cat("Total genes measured:", nrow(data), "\n\n")

# View first few rows
head(data)

```

**💡 CHECKPOINT:** You should see data for 32,842 genes.

---

# SUMMARY STATISTICS

```{r summary-stats}
# Summary of lung cancer data
cat("\n=== LUNG CANCER DATA SUMMARY ===\n")
summary(data[, c("log2FC", "padj")])

```

---

# IDENTIFY DEGs

**The Logic:**
We want genes that are:
1. **Strongly changed**: |log2FC| > 1 (at least 2-fold change)
2. **Statistically significant**: padj < 0.05

**Why these cutoffs?**
- log2FC > 1: Biologically meaningful change (2-fold = 100% change)
- padj < 0.05: 5% chance this is a false positive

```{r identify-degs}
# Define cutoffs
log2fc.cutoff <- 1
pvalue.cutoff <- 0.05

# Filter for DEGs in lung cancer
# abs() = absolute value, captures both up AND down regulated genes
degs <- data[abs(data$log2FC) > log2fc.cutoff & 
             data$padj < pvalue.cutoff, ]

# Separate up and down regulated
degs.up <- degs[degs$log2FC > 0, ]
degs.down <- degs[degs$log2FC < 0, ]

# Print results
cat("\n=== DIFFERENTIALLY EXPRESSED GENES (Lung Cancer) ===\n")
cat("Total DEGs:", nrow(degs), "\n")
cat("  ↑ Up-regulated (higher in lung cancer):", nrow(degs.up), "\n")
cat("  ↓ Down-regulated (lower in lung cancer):", nrow(degs.down), "\n")

# Show top 10 most up-regulated genes
print(degs.up[order(-degs.up$log2FC), ][1:10, c("GeneName", "log2FC", "padj")], row.names = FALSE)

# Show top 10 most down-regulated genes
print(degs.down[order(degs.down$log2FC), ][1:10, c("GeneName", "log2FC", "padj")], row.names = FALSE)

# Save DEG list
write.table(degs, file = paste0(out.folder, "degs-SCLC.tsv"),
            row.names = FALSE, sep = "\t", quote = FALSE)

cat("\n✓ DEG list saved to:", paste0(out.folder, "degs-SCLC.tsv\n"))
```

**💡 CHECKPOINT:** You should find ~6,600 DEGs (lung cancer involves widespread strong changes).

**Look at the top genes:**
- Do you recognize any? 
- Are there more up or down-regulated genes?

---

# VISUALIZE WITH VOLCANO PLOT

**What is a volcano plot?**
- **X-axis**: log2FC (how much gene changed)
  - Left = down in lung cancer
  - Right = up in lung cancer
- **Y-axis**: -log10(p-value) (statistical significance)
  - Higher = more significant
- **Colors**:
  - Red = significantly changed 
  - Blue = significantly but too low log2FC
  - Green = High log2FC but not significant
  - Grey = not significant with low log2FC

```{r volcano-plot}
# Calculate axis limits
max.log2FC <- round(max(abs(data$log2FC), na.rm = TRUE))

# Create volcano plot
volcano <- EnhancedVolcano(
  data,
  title = paste0("Lung cancer vs. healthy (", nrow(degs), " DEGs)"),
  subtitle = paste0("Up-regulated: ", nrow(degs.up), " | Down-regulated: ", nrow(degs.down)),
  lab = data$GeneName,
  x = "log2FC",
  y = "padj",
  pCutoff = pvalue.cutoff,
  FCcutoff = log2fc.cutoff,
  pointSize = 2,
  labSize = 4,
  xlim = c(-max.log2FC, max.log2FC),
  ylim = c(0, 15),
  legendPosition = "right"
)

print(volcano)

# Save high-quality version
png(paste0(out.folder, "volcano-plot-SCLC.png"), 
    width = 2400, height = 1800, res = 200)
print(volcano)
dev.off()

cat("\n✓ Volcano plot saved to:", paste0(out.folder, "volcano-plot-SCLC.png\n"))
```

**💡 LOOK AT THE PLOT:**
- Which side has more DEGs (up or down)?
- Any genes with extreme changes?

---

# SAVE WORKSPACE

```{r save-workspace}
# Save all variables for next module
save.image(file = paste0(out.folder, "module1-workspace.RData"))
cat("✓ Workspace saved for Module 2\n")
```

---

# DISCUSSION QUESTIONS

1. **How many DEGs did you find?**

2. **Up-regulated vs. down-regulated genes**

3. **Look at the top DEGs**
   - Any familiar genes?
   - Do the changes make biological sense?

---

# KEY TAKEAWAYS

✓ We identified ~6,600 DEGs in lung cancer 
✓ We have more up- than down-regulated genes
✓ Volcano plot shows both magnitude and significance  
✓ Lung cancer involves coordinated gene expression changes  

**NEXT MODULE:** We'll find which biological pathways these DEGs belong to.

---