---
title: "MODULE 2: What Pathways? - Pathway Enrichment in Lung Cancer"
author: "mKutmon"
date: "December 2025"
output: html_document
---

# 🎯 Module Goal
Identify which biological pathways contain more DEGs than expected by chance in lung cancer.

---

# SETUP

## Load Workspace from Module 1

```{r load-workspace, message=FALSE, warning=FALSE}
# Load packages
library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)
library(ggplot2)

# Load saved data from Module 1
load("output/module1-workspace.RData")

cat("✓ Loaded from Module 1:\n")
cat("  - Total genes:", nrow(data), "\n")
cat("  - DEGs (lung cancer):", nrow(degs), "\n")
cat("  - Output folder:", out.folder, "\n")
```

**💡 CHECKPOINT:** You should see ~6,600 DEGs from Module 1.

---

# PATHWAY ENRICHMENT ANALYSIS

## The Concept

**Think of it like this:**
- You have a jar with 15,000 marbles (all measured genes)
- 4,000 are RED (your DEGs)
- 11,000 are WHITE (non-DEGs)
- You randomly draw 100 marbles (genes of a specific pathway)
- **Question**: "How many RED marbles did we get? Is it more than expected by chance?"

**Expected by chance:** 4000/15000 × 100 ≈ 27 genes  
**If we observe:** 60 genes → **Enriched!** (p-value tests significance)

**What the statistics mean:**
- **GeneRatio**: "60/4000" = 60 DEGs in pathway / 4000 DEGs in dataset
- **BgRatio**: "100/15000" = 100 pathway genes / 15000 measured genes
- **pvalue**: Probability of seeing this by chance
- **p.adjust**: Corrected for testing hundreds of pathways (FDR method)
- **Count**: Number of our DEGs in this pathway

```{r enrichment-analysis, message=FALSE, warning=FALSE}
# Perform enrichment using WikiPathways database
# WikiPathways is:
# - Community-curated pathway database
# - High quality, regularly updated
# - Covers diverse biological processes
# - Well-integrated with Cytoscape

enrichment <- clusterProfiler::enrichWP(
    gene = degs$GeneID,         # Our DEG list (Entrez IDs)
  organism = "Homo sapiens",    # Human
  pAdjustMethod = "fdr",        # False discovery rate correction
  pvalueCutoff = 0.05           # Show pathways with p.adjust < 0.05
)

# Convert to data frame
enrichment.df <- as.data.frame(enrichment)

# Print summary
cat("\n=== ENRICHMENT SUMMARY ===\n")
cat("Number of enriched pathways:", nrow(enrichment.df), "\n\n")

# Show enriched pathways
print(enrichment.df[, c("Description", "Count", "GeneRatio", "pvalue", "p.adjust")], row.names = FALSE)


# Save full results
write.table(enrichment.df, 
            file = paste0(out.folder, "enrichment-SCLC.tsv"),
            row.names = FALSE, sep = "\t", quote = FALSE)

cat("\n✓ Results saved to:", paste0(out.folder, "enrichment-SCLC.tsv\n"))
```
---

# VISUALIZE ENRICHMENT

## Dot Plot

**What this shows:**
- **Y-axis**: Pathway names (top pathways)
- **X-axis**: Gene ratio (proportion of pathway genes that are DEGs)
- **Dot size**: Number of DEGs in pathway
- **Dot color**: Significance (redder = more significant)

```{r dotplot}
# Create dot plot showing top 50 pathways
if(nrow(enrichment.df) > 0) {
  dotplot(enrichment, showCategory = min(50, nrow(enrichment.df))) + 
    ggtitle("Top Enriched Pathways in Lung Cancer") +
    theme(axis.text.y = element_text(size = 10))
  
  # Save
  ggsave(paste0(out.folder, "enrichment-dotplot-SCLC.png"), 
         width = 10, height = 15, dpi = 200)
  
  cat("✓ Dot plot saved to:", paste0(out.folder, "enrichment-dotplot-SCLC.png\n"))
} else {
  cat("⚠ No enriched pathways to plot\n")
}
```

---

## Tree Plot

**Alternative visualization:**
Shows the relationship (overlapping genes) between the enriched pathways. This is very helpful for interpretation.

```{r treeplot}
# Create bar plot
if(nrow(enrichment.df) > 0) {
  # Create treeplot to show similarity between pathways (overlapping genes)
  enrichment.sim <- enrichplot::pairwise_termsim(enrichment)
  treeplot(enrichment.sim, showCategory = 50, cluster.params=list(n = 10, label_words_n = 0))
  
  # Save
  ggsave(paste0(out.folder, "enrichment-treeplot-SCLC.png"), 
         width = 20, height = 15, dpi = 200)
  
  cat("✓ Treeplot plot saved to:", paste0(out.folder, "enrichment-treeplot-SCLC.png\n"))
}
```
---

# SAVE WORKSPACE

```{r save-workspace}
# Save all variables for next module
save.image(file = paste0(out.folder, "module2-workspace.RData"))
cat("✓ Workspace saved for Module 3\n")
```

---

# DISCUSSION QUESTIONS (5 min)

**For the instructor to discuss with students:**

1. **What are the top enriched pathways?**

2. **Look at the Gene Ratio values**
   - Which pathway has the highest proportion of DEGs?
   - What does a GeneRatio of "60/100" mean?

---

# KEY TAKEAWAYS

✓ Pathway enrichment reveals biological context of DEGs  
✓ Many pathways are affected in lung cancer
✓ Cell cycle, immune response, complement system, differentiation, etc.
✓ Statistical correction (FDR) prevents false discoveries  
✓ Gene Ratio shows what proportion of pathway is affected  

**NEXT MODULE:** We'll visualize these pathways in Cytoscape with our expression data.
