---
title: "OPTIONAL EXTENSIONS: Network Analysis & Drug Targets in lung cancer"
author: "mkutmon"
date: "December 2025"
output: html_document
---

# 🎯 Optional Extensions

**If you have extra time**, explore these advanced topics:
1. Network view of pathways
2. Hub node identification  
3. Drug-target extension 

---

# EXTENSION 1: Network View of Pathway

## Concept

**Pathway diagram vs. Network:**
- **Pathway diagram** = Shows biological layout (textbook view)
- **Network** = Shows connectivity (nodes = genes, edges = interactions)

Network view helps identify:
- Highly connected genes (hubs)
- Interaction patterns
- Functional modules

---

## Setup

```{r setup-ext1, message=FALSE, warning=FALSE}
library(RCy3)
library(RColorBrewer)

# Load workspace
load("output/module2-workspace.RData")

# Make sure Cytoscape is running
cytoscapePing()

# Get get the Cell Cycle pathway again
# Note: Not all pathways have a good network structure
pathway.id <- "WP179"
pathway.name <- enrichment.df$Description[enrichment.df$ID == pathway.id]

cat("Working with pathway:", pathway.name, "\n")
```

---

## Import Pathway as Network

**Key difference:** 
Using `import-as-network` instead of `import-as-pathway`

```{r import-network}
# Import as network (not diagram)
cat("Importing pathway as network...\n")

RCy3::commandsRun(paste0('wikipathways import-as-network id=', pathway.id))
Sys.sleep(3)

RCy3::mapTableColumn(column = "Ensembl", species = "human", map.from = "Ensembl", map.to = "Entrez Gene")


# Get network info
network.name <- RCy3::getNetworkName()
node.count <- RCy3::getNodeCount()
edge.count <- RCy3::getEdgeCount()

cat("\n=== NETWORK STATISTICS ===\n")
cat("Network name:", network.name, "\n")
cat("Nodes (genes/proteins):", node.count, "\n")
cat("Edges (interactions):", edge.count, "\n")
```

---

## Apply Expression Data

```{r apply-data-network}
# Load expression data
RCy3::loadTableData(data, data.key.column = "GeneID", table.key.column = "Entrez Gene")

# Apply color mapping
data.values <- c(-1, 0, 1)
node.colors <- c("#4393C3", "#FFFFFF", "#D6604D")  # Blue, White, Red

RCy3::setNodeColorMapping(
  table.column = "log2FC",
  table.column.values = data.values,
  colors = node.colors,
  mapping.type = "c",
  default.color = "#CCCCCC",
  style.name = "WikiPathways-As-Network"
)

# Add significance borders
RCy3::setNodeBorderColorMapping(
  table.column = "padj",
  table.column.values = c(0, 0.05),
  colors = c("#00FF00", "#CCCCCC"),
  mapping.type = "c",
  style.name = "WikiPathways-As-Network"
)

RCy3::setNodeBorderWidthMapping(
  table.column = "padj",
  table.column.values = c(0, 0.05, 1),
  widths = c(8, 8, 1),
  mapping.type = "c",
  style.name = "WikiPathways-As-Network"
)

cat("✓ Expression data applied to network\n")
```

---

## Apply Layout

Network layouts help visualize connectivity patterns.

```{r apply-layout}
# Try force-directed layout (groups connected nodes together)
cat("Applying layout...\n")
RCy3::layoutNetwork("force-directed")

cat("✓ Layout applied\n")
cat("\n🔍 Look at Cytoscape: Network view shows connectivity!\n")
```

---

## Export Network View

```{r export-network}
RCy3::exportImage(
  filename = paste0(out.folder, pathway.id , "-network.png"),
  type = "PNG",
  zoom = 200
)

cat("✓ Network view saved to:", paste0(out.folder, pathway.id, "-network.png\n"))
```

**Compare:**
- Pathway diagram vs. network view
- Which is easier to see connectivity?
- Which shows biological context better?

---

# EXTENSION 2: Hub Node Identification

## Concept

**Hub nodes** = Genes with many connections (interactions)

**Why care about hubs in lung cancer?**
- Often essential
- Master regulators
- Potentially good therapeutic targets (e.g., p53, p21)
- Dysregulation has cascading effects

---

## Calculate Node Degree

**Degree** = Number of connections a node has

```{r calculate-degree}
# Calculate degree for all nodes
# This counts how many edges connect to each node
RCy3::analyzeNetwork()

cat("✓ Network analysis complete\n")
cat("\n🔍 In Cytoscape: Check Node Table - new 'Degree' column added\n")
```

---

## Identify Top Hubs

```{r identify-hubs}
# Get node table with degree information
node.table <- RCy3::getTableColumns(table = "node")

# Filter for nodes with expression data
nodes.with.data <- node.table[!is.na(node.table$log2FC), ]

# Find top 10 most connected genes
if("Degree" %in% colnames(nodes.with.data)) {
  top.hubs <- nodes.with.data[order(-nodes.with.data$Degree), ]
  top.hubs <- top.hubs[1:min(10, nrow(top.hubs)), 
                       c("name", "Degree", "log2FC", "padj")]
  
  cat("\n=== TOP 10 HUB GENES IN  PATHWAY ===\n")
  print(top.hubs, row.names = FALSE)
  
  # Save
  write.table(top.hubs,
              file = paste0(out.folder, "hub-genes-SCLC.tsv"),
              row.names = FALSE, sep = "\t", quote = FALSE)
  
  cat("\n✓ Hub genes saved\n")
} else {
  cat("⚠ Degree column not found. Make sure network analyzer ran successfully.\n")
}
```

**Questions:**
- Are hub genes also DEGs in lung cancer?
- Do hubs tend to be up or down-regulated?

---

## Visualize Hubs

Make hub nodes larger so they stand out.

```{r visualize-hubs}
# Map node size to degree (more connections = larger node)
if("Degree" %in% colnames(node.table)) {
  RCy3::setNodeSizeMapping(
    table.column = "Degree",
    table.column.values = c(1, 5, 10, 20),
    sizes = c(30, 50, 70, 100),
    mapping.type = "c",
    style.name = "WikiPathways-As-Network"
  )
  
  RCy3::selectNodes(nodes = "Group", by.col = "Type")
  RCy3::setNodeSizeBypass(RCy3::getSelectedNodes(),new.sizes = 20)
  RCy3::clearSelection()
  cat("✓ Node sizes mapped to degree\n")
  cat("\n🔍 In Cytoscape: Hub genes are now larger!\n")
  
  # Export
  RCy3::exportImage(
    filename = paste0(out.folder, pathway.id, "-network-hubs-SCLC.png"),
    type = "PNG",
    zoom = 200
  )
  
  cat("✓ Hub visualization saved\n")
}
```

---

# EXTENSION 3: Drug-Target Extension (Senolytics!)

## Concept

**Finding therapeutic opportunities:** Now that we've identified a dysregulated pathway, we can extend it with known drug-target interactions. This helps identify: - **Existing drugs** that target
genes in this pathway - **Potential repurposing opportunities** for cancer treatment - **Mechanistic understanding** of how drugs might work

This section uses the **CyTargetLinker** app to add drug information to our pathway network.

---

## Install CyTargetLinker

```{r install-cytargetlinker}
# Check if installed
status <- RCy3::getAppStatus("CyTargetLinker")

if (!grepl("status: Installed", status)) {
  cat("Installing CyTargetLinker app...\n")
  RCy3::installApp("CyTargetLinker")
  Sys.sleep(5)
  cat("✓ CyTargetLinker installed\n")
} else {
  cat("✓ CyTargetLinker already installed\n")
}
```

---

## Extend Network with Drugs

**Using Cytoscape GUI (Recommended)**

```{r eval=FALSE}

unzip(system.file("extdata","drugbank-5.1.0.xgmml.zip", package="rWikiPathways"), exdir = getwd())
drugbank.file <- file.path(getwd(), "drugbank-5.1.0.xgmml")

# Extend network with drug-target interactions
RCy3::commandsRun(paste0('cytargetlinker extend idAttribute="Ensembl" linkSetFiles="', drugbank.file, '"'))

# Apply automatic layout
RCy3::commandsRun('cytargetlinker applyLayout network="current"')

# Style drug nodes for better visibility
drug.nodes <- RCy3::selectNodes("drug", by.col = "CTL.Type", preserve = FALSE)$nodes
RCy3::clearSelection()
RCy3::setNodeColorBypass(drug.nodes, "#DD99FF")  # Purple color
RCy3::setNodeShapeBypass(drug.nodes, "hexagon")   # Hexagon shape

# Get drug labels and make them more visible
drug.data <- RCy3::getTableColumns(columns = c("SUID", "CTL.label"))
drug.labels <- drug.data[!is.na(drug.data$CTL.label), ]
if(nrow(drug.labels) > 0) {
  RCy3::setNodeLabelBypass(drug.labels$SUID, drug.labels$CTL.label)
}

# Export the extended network
RCy3::exportImage(
  filename = paste0(out.folder, pathway.id, "-with-drugs.png"),
  type = "PNG",
  zoom = 500
)

cat("\nDrug-extended network created!\n")
cat("Purple hexagons = drugs that target genes in this pathway\n")
```

---

# SUMMARY OF EXTENSIONS

## What We Learned

**Extension 1: Network View**
✓ Networks reveal connectivity patterns  
✓ Complementary to pathway diagrams  
✓ Helps identify interaction modules  

**Extension 2: Hub Nodes**
✓ Hubs = highly connected, often regulatory genes  
✓ Potentially good candidates for intervention  

**Extension 3: Drug Targets**
✓ CyTargetLinker connects pathways to drugs  
✓ Identifies potential drug candidates for intervention
