---
title: "MODULE 3: See It! - Pathway Visualization in Cytoscape"
author: "mkutmon"
date: "December 2025"
output: html_document
---

# 🎯 Module Goal
Visualize gene expression changes in senescence on biological pathway diagrams using Cytoscape.

---

# SETUP

## Load Workspace from Module 2

```{r load-workspace, message=FALSE, warning=FALSE}
# Load packages
library(RCy3)
library(RColorBrewer)

# Load saved data from Module 2
load("output/module2-workspace.RData")

cat("✓ Loaded from Module 2:\n")
cat("  - Total genes:", nrow(data), "\n")
cat("  - DEGs:", nrow(degs), "\n")
cat("  - Enriched pathways:", nrow(enrichment.df), "\n")
```

---

## IMPORTANT: Open Cytoscape!

**Before running the code below:**
1. Open the Cytoscape application
2. Wait for it to fully load (may take 30-60 seconds)
3. Keep it open in the background

---

## Test Connection to Cytoscape

```{r test-connection}
# Try to connect to Cytoscape
tryCatch({
  RCy3::cytoscapePing()
  cat("✓ Successfully connected to Cytoscape!\n")
}, error = function(e) {
  cat("✗ Could not connect to Cytoscape\n")
  cat("  Make sure Cytoscape is running and try again.\n")
  cat("  Error:", e$message, "\n")
})
```

**💡 CHECKPOINT:** You should see "Successfully connected to Cytoscape!"

**If connection fails:**
1. Make sure Cytoscape is actually open
2. Restart Cytoscape
3. Re-run the code chunk above

---

# VISUALIZE TOP ENRICHED PATHWAY

## Choose Pathway

Let's visualize the top enriched pathway from our analysis.

```{r choose-pathway}
# Get top pathway
if(nrow(enrichment.df) > 0) {
  pathway.id <- enrichment.df$ID[1]
  pathway.name <- enrichment.df$Description[1]
  pathway.count <- enrichment.df$Count[1]
  
  cat("=== PATHWAY TO VISUALIZE ===\n")
  cat("Name:", pathway.name, "\n")
  cat("ID:", pathway.id, "\n")
  cat("DEGs in pathway:", pathway.count, "\n")
} else {
  cat("⚠ No enriched pathways found\n")
}
```

---

## Install WikiPathways App

Cytoscape needs the WikiPathways app to import pathways.

```{r install-app}
# Check if WikiPathways app is installed
status <- RCy3::getAppStatus("wikipathways")

if (!grepl("status: Installed", status)) {
  cat("Installing WikiPathways app...\n")
  RCy3::installApp("wikipathways")
  Sys.sleep(3)  # Wait for installation
  cat("✓ WikiPathways app installed\n")
} else {
  cat("✓ WikiPathways app already installed\n")
}
```

---

## Import Pathway into Cytoscape

**What this does:**
Imports the pathway diagram from WikiPathways database into Cytoscape.

```{r import-pathway}
if(exists("pathway.id")) {
  # Import pathway as a diagram (not as network)
  cat("Importing pathway into Cytoscape...\n")
  
  RCy3::commandsRun(paste0('wikipathways import-as-pathway id=', pathway.id))
  
  # Give Cytoscape time to load
  Sys.sleep(3)
  
  # Toggle graphics details for better view
  RCy3::toggleGraphicsDetails()
  
  cat("✓ Pathway imported successfully!\n")
  cat("\n🔍 Look at Cytoscape - you should see the pathway diagram!\n")
}
```

**💡 CHECKPOINT:** Look at Cytoscape window - you should see a pathway diagram with genes, proteins, and metabolites.

---

## Load Expression Data

**What this does:**
Adds our gene expression data (log2FC, p-values) to the pathway nodes in Cytoscape.

```{r load-data}
# Load expression data into Cytoscape
# Match by Ensembl ID
RCy3::loadTableData(
  data, 
  data.key.column = "GeneID",      # Column in our data
  table.key.column = "Ensembl"     # Column in Cytoscape nodes
)

cat("✓ Expression data loaded into Cytoscape\n")
cat("\n🔍 In Cytoscape: Click 'Node Table' (bottom) to see log2FC and p-value columns (scroll to the right)\n")
```

**💡 CHECKPOINT:** In Cytoscape, look at the Node Table (bottom panel). You should see new columns: log2FC, p.value

---

## Apply Color Mapping

**What this does:**
Colors nodes based on gene expression in senescence:
- **Red** = UP-regulated in senescence (positive log2FC)
- **Blue** = DOWN-regulated in senescence (negative log2FC)  
- **White** = No change (log2FC = 0)

The more intense the color, the stronger the change.

```{r color-mapping}
# Define color scheme
# Map log2FC from -1 (blue) to 0 (white) to +1 (red)
data.values <- c(-2, 0, 2)
node.colors <- c("#4393C3", "#FFFFFF", "#D6604D")  # Blue, White, Red

# Apply color mapping to nodes
RCy3::setNodeColorMapping(
  table.column = "log2FC",    # Use replicative log2FC values
  table.column.values = data.values,      # Define scale
  colors = node.colors,                   # Define colors
  mapping.type = "c",                     # "c" = continuous
  style = "WikiPathways"                  # Apply to WikiPathways style
)

cat("✓ Color mapping applied\n")
cat("\n🔍 In Cytoscape: Nodes should now be colored red/blue based on expression!\n")
```

**💡 INTERPRETATION:**
- **Red nodes** = genes UP in senescence 
- **Blue nodes** = genes DOWN in senescence 
- **White/grey nodes** = no change or no data

---

## Add Significance Borders

**What this does:**
Adds green borders to genes that are statistically significant (p < 0.05).

```{r border-mapping}
# Map border color based on significance
# Green = significant (p < 0.05)
# Grey = not significant
RCy3::setNodeBorderColorMapping(
  table.column = "p.value",
  table.column.values = c(0, 0.05),
  colors = c("#00FF00", "#CCCCCC"),  # Green, Grey
  mapping.type = "c",
  style.name = "WikiPathways"
)

# Make borders thicker for significant genes
RCy3::setNodeBorderWidthMapping(
  table.column = "p.value",
  table.column.values = c(0, 0.05, 1),
  widths = c(5, 5, 0),               # Thick for p<0.05, none otherwise
  mapping.type = "c",
  style.name = "WikiPathways"
)

cat("✓ Significance borders applied\n")
cat("\n🔍 In Cytoscape: Significant genes now have thick green borders!\n")
```

---

## Export Visualization

Save the pathway visualization as an image.

```{r export-image}
if(exists("pathway.id")) {
  # Export as high-quality PNG
  RCy3::exportImage(
    filename = paste0(out.folder, pathway.id, "-pathway-replicative.png"),
    type = "PNG",
    zoom = 500
  )
  
  cat("✓ Pathway visualization saved to:", 
      paste0(out.folder, pathway.id, "-pathway-replicative.png\n"))
}
```

---

# INTERPRET THE VISUALIZATION

**Look at the pathway in Cytoscape and answer:**

1. **Which parts of the pathway are most affected?**
   - Where do you see clusters of red or blue?
   - Are entire branches affected?

2. **Color patterns:**
   - More red (up) or blue (down)?
   - Mixed regulation or coordinated?

3. **Do the changes make biological sense?**
   - For senescence, what would you expect?
   - Cell cycle genes? DNA damage genes?

4. **Significant genes (green borders):**
   - Are most changed genes also significant?
   - Any highly significant but small changes?
   
---

# TRY ANOTHER PATHWAY (HANDS-ON)

**Your task:** Choose a different pathway and visualize it!

```{r student-activity}
# List available enriched pathways
if(nrow(enrichment.df) > 0) {
  cat("=== AVAILABLE PATHWAYS TO VISUALIZE ===\n")
  cat("Choose one by row number:\n\n")
  
  n.pathways <- min(10, nrow(enrichment.df))
  pathway.options <- enrichment.df[1:n.pathways, c("Description", "Count", "p.adjust")]
  pathway.options$Row <- 1:n.pathways
  print(pathway.options[, c("Row", "Description", "Count", "p.adjust")], row.names = FALSE)
  
  # TASK: Change this number to choose a pathway (1-10)
  chosen.row <- 7  # <-- CHANGE THIS NUMBER
  
  # Get pathway info
  student.pathway.id <- enrichment.df$ID[chosen.row]
  student.pathway.name <- enrichment.df$Description[chosen.row]
  
  cat("\n=== YOU CHOSE ===\n")
  cat("Pathway:", student.pathway.name, "\n")
  cat("ID:", student.pathway.id, "\n")
}
```

**Now run the visualization:**

```{r student-visualization}
if(exists("student.pathway.id")) {
  # Import pathway
  RCy3::commandsRun(paste0('wikipathways import-as-pathway id=', student.pathway.id))
  Sys.sleep(3)
  RCy3::toggleGraphicsDetails()
  
  # Load data
  RCy3::loadTableData(data, data.key.column = "GeneID", table.key.column = "Ensembl")
  
  # Apply colors (same as before)
  data.values <- c(-1, 0, 1)
  node.colors <- c("#4393C3", "#FFFFFF", "#D6604D")  # Blue, White, Red
  
  RCy3::setNodeColorMapping(
    table.column = "log2FC",
    table.column.values = data.values,
    colors = node.colors,
    mapping.type = "c",
    style = "WikiPathways"
  )
  
  # Add borders
  RCy3::setNodeBorderColorMapping(
    table.column = "p.value",
    table.column.values = c(0, 0.05),
    colors = c("#00FF00", "#CCCCCC"),
    mapping.type = "c",
    style.name = "WikiPathways"
  )
  
  RCy3::setNodeBorderWidthMapping(
    table.column = "p.value",
    table.column.values = c(0, 0.05, 1),
    widths = c(5, 5, 0),
    mapping.type = "c",
    style.name = "WikiPathways"
  )
  
  # Export
  RCy3::exportImage(
    filename = paste0(out.folder, student.pathway.id, "-pathway-student.png"),
    type = "PNG",
    zoom = 500
  )
  
  cat("\n✓ Your pathway saved to:", 
      paste0(out.folder, student.pathway.id, "-pathway-student.png\n"))
}
```

**Compare with your neighbor:**
- Which pathway did you choose?
- What patterns do you see?
- Are the changes what you expected?
- How does it compare to the first pathway?

---

# SAVE SESSION

```{r save-session}
# Save Cytoscape session for later
# This preserves all networks, styles, and data
session.file <- paste0(out.folder, "cytoscape-session-senescence.cys")

RCy3::saveSession(session.file)

cat("✓ Cytoscape session saved to:", session.file, "\n")
cat("  You can reopen this later in Cytoscape: File → Open → Session\n")
```

---

# KEY TAKEAWAYS

✓ Pathway diagrams show molecular context of gene expression changes  
✓ Color coding reveals spatial patterns and coordinated regulation  
✓ Red = up in senescence, Blue = down, Green border = significant  
✓ Senescence pathways show characteristic patterns:
  - SASP activation (cytokines up)
  - Cell cycle arrest (CDK inhibitors up, cyclins down)
  - DNA damage response (checkpoint genes up)
✓ Visual inspection reveals biological insights not obvious from gene lists  

**OPTIONAL:** Continue to Extensions for network analysis and drug targets!

---

# SENESCENCE BIOLOGY SUMMARY

**What we've learned:**

✓ **Molecular signature**: Thousands of genes changed  
✓ **Coordinated program**: Multiple pathways affected  
✓ **SASP**: Inflammatory secretome active  
✓ **Cell cycle arrest**: CDK inhibitors up, cyclins down  
✓ **DNA damage**: DDR pathway active  

**Functional consequences:**
- Cells stop dividing (tumor suppression)
- Secrete factors (affect neighbors)
- Resist apoptosis (accumulate with age)
- Impair tissue function (contribute to aging)

