---
title: "MODULE 1: What Changed? - Identifying DEGs in Senescence"
author: "mkutmon"
date: "December 2025"
output: html_document
---

# 🎯 Module Goal
Identify which genes are significantly changed in replicative senescence compared to immortal cells.

---

# SETUP

## Install & Load Packages

```{r packages, message=FALSE, warning=FALSE}
# Load packages
library(EnhancedVolcano)
library(rstudioapi)

cat("✓ Packages loaded successfully!\n")
```

## Set Working Directory

```{r}
# Set working directory to script location
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

# Create output folder
out.folder <- "output/"
if(!dir.exists(out.folder)) dir.create(out.folder)

cat("✓ Working directory:", getwd(), "\n")
cat("✓ Output folder created\n")
```

---

# LOAD DATA

**What's in this dataset:**
The dataset contains gene expression measurements comparing cellular senescence with immortal (non-senescent) cells.

**Columns:**
- **Ensembl**: Gene identifier
- **Symbol**: Gene name (e.g., CDKN1A, TP53)
- **Entrezid**: Alternative gene ID
- **log2FC**: Log2 fold change (replicative senescent vs. immortal)
  - **Positive** = UP in senescence
  - **Negative** = DOWN in senescence
- **p-val**: P-value for this comparison

**Biological context:**
- **Replicative senescence** = Natural aging after many cell divisions
- **Immortal cells** = Can divide indefinitely (baseline)
- Senescence = permanent cell cycle arrest, typically a tumor suppressor mechanism

```{r load-data}
# Load cell senescence dataset
data <- read.csv("dataset-cell-senescene.csv")

# Quick check
cat("Total genes measured:", nrow(data), "\n\n")

# View first few rows
head(data)

```

**💡 CHECKPOINT:** You should see data for 36,608 genes.

---

# SUMMARY STATISTICS

```{r summary-stats}
# Summary of replicative senescence data
cat("\n=== REPLICATIVE SENESCENCE DATA SUMMARY ===\n")
summary(data[, c("log2FC", "p.value")])

```

---

# IDENTIFY DEGs

**The Logic:**
We want genes that are:
1. **Strongly changed**: |log2FC| > 1 (at least 2-fold change)
2. **Statistically significant**: p-value < 0.05

**Why these cutoffs?**
- log2FC > 1: Biologically meaningful change (2-fold = 100% change)
- p-value < 0.05: 5% chance this is a false positive
- Note: This dataset uses raw p-values (not adjusted) because fewer comparisons

```{r identify-degs}
# Define cutoffs
log2fc.cutoff <- 1
pvalue.cutoff <- 0.05

# Filter for DEGs in REPLICATIVE senescence
# abs() = absolute value, captures both up AND down regulated genes
degs <- data[abs(data$log2FC) > log2fc.cutoff & 
             data$p.value < pvalue.cutoff, ]

# Separate up and down regulated
degs.up <- degs[degs$log2FC > 0, ]
degs.down <- degs[degs$log2FC < 0, ]

# Print results
cat("\n=== DIFFERENTIALLY EXPRESSED GENES (Replicative Senescence) ===\n")
cat("Total DEGs:", nrow(degs), "\n")
cat("  ↑ Up-regulated (higher in senescence):", nrow(degs.up), "\n")
cat("  ↓ Down-regulated (lower in senescence):", nrow(degs.down), "\n")

# Show top 10 most up-regulated genes
print(degs.up[order(-degs.up$log2FC), ][1:10, c("GeneName", "log2FC", "p.value")], row.names = FALSE)

# Show top 10 most down-regulated genes
print(degs.down[order(degs.down$log2FC), ][1:10, c("GeneName", "log2FC", "p.value")], row.names = FALSE)

# Save DEG list
write.table(degs, file = paste0(out.folder, "degs-replicative.tsv"),
            row.names = FALSE, sep = "\t", quote = FALSE)

cat("\n✓ DEG list saved to:", paste0(out.folder, "degs-replicative.tsv\n"))
```

**💡 CHECKPOINT:** You should find ~4,000 DEGs (senescence involves widespread changes).

**Look at the top genes:**
- Do you recognize any? 
- Are there more up or down-regulated genes?

---

# VISUALIZE WITH VOLCANO PLOT

**What is a volcano plot?**
- **X-axis**: log2FC (how much gene changed)
  - Left = down in senescence
  - Right = up in senescence
- **Y-axis**: -log10(p-value) (statistical significance)
  - Higher = more significant
- **Colors**:
  - Red = significantly changed 
  - Blue = significantly but too low log2FC
  - Green = High log2FC but not significant
  - Grey = not significant with low log2FC

```{r volcano-plot}
# Calculate axis limits
max.p.val <- round(max(-log10(data$p.value), na.rm = TRUE))
max.log2FC <- round(max(abs(data$log2FC), na.rm = TRUE))

# Create volcano plot
volcano <- EnhancedVolcano(
  data,
  title = paste0("Replicative Senescence vs. Immortal Cells (", nrow(degs), " DEGs)"),
  subtitle = paste0("Up-regulated: ", nrow(degs.up), " | Down-regulated: ", nrow(degs.down)),
  lab = data$GeneName,
  x = "log2FC",
  y = "p.value",
  pCutoff = pvalue.cutoff,
  FCcutoff = log2fc.cutoff,
  pointSize = 2,
  labSize = 4,
  xlim = c(-max.log2FC, max.log2FC),
  ylim = c(0, max.p.val),
  legendPosition = "right"
)

print(volcano)

# Save high-quality version
png(paste0(out.folder, "volcano-plot-replicative.png"), 
    width = 2400, height = 1800, res = 200)
print(volcano)
dev.off()

cat("\n✓ Volcano plot saved to:", paste0(out.folder, "volcano-plot-replicative.png\n"))
```

**💡 LOOK AT THE PLOT:**
- Which side has more DEGs (up or down)?
- Any genes with extreme changes?

---

# BIOLOGICAL INTERPRETATION

```{r interpretation}
cat("\n=== BIOLOGICAL PATTERNS ===\n\n")

# Are more genes up or down?
if(nrow(degs.up) > nrow(degs.down)) {
  cat("✓ More genes are UP-regulated in senescence\n")
  cat("  This suggests active transcriptional programs\n")
  cat("  SASP (senescence-associated secretory phenotype)\n\n")
} else {
  cat("✓ More genes are DOWN-regulated in senescence\n")
  cat("  This suggests repression of proliferation programs\n\n")
}

# Ratio of up to down
ratio <- round(nrow(degs.up) / nrow(degs.down), 2)
cat("Ratio of up/down-regulated:", ratio, "\n")
```

---

# SAVE WORKSPACE

```{r save-workspace}
# Save all variables for next module
save.image(file = paste0(out.folder, "module1-workspace.RData"))
cat("✓ Workspace saved for Module 2\n")
```

---

# DISCUSSION QUESTIONS (7 min)

**For the instructor to discuss with students:**

1. **How many DEGs did you find?**

2. **Up-regulated vs. down-regulated genes**

3. **Look at the top DEGs and senescence markers**
   - Any familiar genes?
   - Do the changes make biological sense?

---

# KEY TAKEAWAYS

✓ We identified ~4,000 DEGs in replicative senescence  
✓ More genes are up-regulated than down-regulated  
✓ Volcano plot shows both magnitude and significance  
✓ Senescence involves coordinated gene expression changes  

**NEXT MODULE:** We'll find which biological pathways these DEGs belong to.

---